---
title: "PLToolkit walkthrough: NegKTSwt (WT1(-)KTS, wild-type) Ind vs Mock Ind"
author: "Xin Wang"
date: "`r format(Sys.Date())`"
description: >
  A step-by-step, narrated walkthrough of Examples/example_NegKTSwt_vs_Mock.R -
  every step the PLToolkit pipeline performs for the NegKTSwt-vs-Mock proximity
  labeling comparison, with the statistical reasoning behind each step spelled
  out inline. Uses wild-type WT1(-)KTS as the bait, unlike
  example_R467W_vs_Mock.R (a pathogenic/deficient variant) - preferred as the
  toolkit's flagship walkthrough since it reflects normal WT1 function.
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)
```

## What this comparison is

`NegKTSwt` is the wild-type WT1(-)KTS isoform (the "Ind" suffix means the
bait's expression was induced), proximity-labeled and pulled down against a
`Mock` (no-bait) induced control. Every protein identified in this
experiment is either genuine WT1(-)KTS proximal-interactome signal or one of
the noise sources proximity labeling is prone to (stochastic background
biotinylation, streptavidin-pulldown contaminants, missing-not-at-random
low-abundance dropout) - the entire pipeline below exists to separate the
two as rigorously as the data allows.

## Step 1 - Source the toolkit

```{r source-toolkit}
toolkit_dir <- "/Users/XXW004/Documents/Projects/MannNina/Project/WT1/scripts/PLToolkit"

for (f in list.files(file.path(toolkit_dir, "R"), pattern = "\\.R$", full.names = TRUE)) {
  source(f)
}
```

This sources every `R/*.R` file in the toolkit - there's no package/DESCRIPTION,
so nothing else is needed beyond the CRAN/Bioconductor dependencies listed in
the toolkit's README (`limma`, `DEqMS`, `DEP`, `proDA`, `MSstats`, `ROTS`,
`impute`, `imputeLCMD`, `metap`, `qvalue`, ...).

## Step 2 - Define the comparison (`comparison_spec`)

```{r spec}
spec <- comparison_spec(
  name = "NegKTSwt_vs_Mock",
  group1_name = "NegKTSwt_Ind", group1_cols = paste0("NegKTSwt_Ind_", 1:3),
  group2_name = "Mock_Ind", group2_cols = paste0("Mock_Ind_", 1:3),
  max_na = 3
)
```

`comparison_spec()` is a plain list bundling everything downstream needs to
know about *this one* bait-vs-control comparison: which sample columns belong
to the bait group (`group1_cols`) and which to the control (`group2_cols`),
human-readable names for each, and `max_na` - the maximum number of missing
values a protein is allowed to have across these 6 samples before it's
dropped entirely (see Step 4). Nothing about *how* the comparison is analyzed
lives here - that's all downstream in `run_pl_dea()`.

## Step 3 - Point at the input files

```{r paths}
paths <- list(
  combined_protein_path = "/Users/XXW004/Documents/Projects/MannNina/Project/WT1/Preproceed/combined_protein.csv",
  psm_dir = "/Users/XXW004/Library/CloudStorage/OneDrive-NationwideChildren'sHospital/NinaMannLab/WT1/Preproceed/",
  msstats_path = "/Users/XXW004/Documents/Projects/MannNina/Project/WT1/Preproceed/MSstats.csv",
  combined_peptide_path = "/Users/XXW004/Documents/Projects/MannNina/Project/WT1/Preproceed/combined_peptide.tsv",
  contaminant_list_path = NULL,
  out_dir = "/Users/XXW004/Documents/Projects/MannNina/Project/WT1/Results/NegKTSwtvsMockInd/PLToolkitDemo/"
)
```

Only `combined_protein_path` and `out_dir` are strictly required. The rest are
optional and each unlocks one extra piece of the analysis if supplied:

- **`psm_dir`** - PSM-level identification counts, needed for DEqMS's
  per-protein quantification-depth covariate (Step 5c). Without it, DEqMS is
  skipped.
- **`msstats_path`** - FragPipe's separate feature-level MSstats export.
  Without it, the MSstats method is skipped.
- **`combined_peptide_path`** - peptide-level identification counts, used
  purely as a QC annotation (`min_peptides_per_sample`) flagging proteins
  whose MaxLFQ intensity rests on very few peptides in some sample - doesn't
  filter anything, just tells you which DEP calls to sanity-check by eye.
- **`contaminant_list_path`** - a CRAPome-style contaminant list for
  filtering common streptavidin-pulldown contaminants (keratins,
  carboxylases). `NULL` here means no contaminant filtering is applied in
  this run.

## Step 4 - Run the pipeline (`run_pl_dea`)

```{r run-pipeline}
out <- run_pl_dea(spec, paths)
```

This one call does the entire intensity-branch analysis. Unpacked:

### 4a. Load and extract

`load_fragpipe_protein()` reads `combined_protein.csv`, restricts to the
target organism, dedupes by `Gene`, and strips FragPipe's `_HUMAN` suffixes.
`extract_intensity_matrix()` then pulls out this comparison's 6 MaxLFQ
intensity columns, recodes zero (not detected) as `NA`, and drops any
protein with more than `spec$max_na` missing values across those 6 samples.

### 4b. Normalize

`log2_quantile_normalize()` log2-transforms the retained intensities and
quantile-normalizes across samples (`limma::normalizeBetweenArrays`) -
correcting for the bait/enzyme-activity-driven intensity shifts that differ
between replicates even before biology is considered.

### 4c. Classify and impute missingness

Not all missingness means the same thing. `classify_mnar()` flags a protein
as **MNAR** (missing not at random) if it's entirely absent in one whole
condition group - consistent with truly being below the limit of detection
in that condition, not random dropout. Everything else is **MAR** (missing
at random, scattered dropouts). `impute_mixed()` then imputes each class
differently: k-nearest-neighbors (`impute::impute.knn`) for MAR values, and
left-censored minimum-probability imputation (`imputeLCMD::impute.MinProb`,
quantile 0.01) for MNAR values - applied directly to the full matrix rather
than through `DEP::impute(fun="mixed")`, which was verified on this data to
scramble the sample/protein orientation MinProb expects for a modest MNAR
class size, producing implausibly high "low-abundance" draws.

The fully-imputed, normalized log2 matrix is written out here too, at
`out_dir/NegKTSwt_vs_Mock__normalized_imputed_matrix.csv`, and is what four
of the seven DEA methods below actually test.

### 4d. Seven independent DEA methods

| Method | What it does differently |
| --- | --- |
| **limma** | Moderated t-test: empirical-Bayes shrinkage of each protein's variance toward a common prior, on the shared imputed matrix. |
| **DEqMS** | Same limma fit, but the empirical-Bayes prior variance is additionally a function of each protein's minimum PSM count (from `psm_dir`) - corrects for proteins quantified from very few PSMs having systematically noisier variance than limma alone assumes. |
| **Wilcoxon** | Non-parametric rank-sum test per protein, same imputed matrix - no distributional assumption, so a useful check against the parametric methods. |
| **ROTS** | A modified t-statistic whose regularization is chosen by optimizing reproducibility of the top gene list across bootstrap resamples (10,000 resamples, optimized over the top 500 genes). |
| **DEP package** | A fully separate, self-contained pipeline on the *raw* (non-shared, non-quantile-normalized) intensities: its own vsn normalization, its own KNN imputation, its own moderated test. |
| **proDA** | Doesn't impute at all - models left-censored missingness directly in its likelihood. Runs on the log2-only (not quantile-normalized, not imputed) matrix with its own median normalization. |
| **MSstats** | Entirely independent of `combined_protein.csv` - runs on FragPipe's separate feature-level export, with its own summarization, normalization, and linear mixed-effects model. |

Every method's output is standardized to the same (Gene, logFC, P.Value,
adj.P.Val) shape so they can be joined and compared apples-to-apples next.

```{r inspect-results}
cat("Proteins after filtering:", nrow(out$intens_raw), "\n")
cat("Methods run:", paste(names(out$results), collapse = ", "), "\n")
```

### 4e. Consensus scoring (`combine_pvalues`, called internally)

A protein must be testable by *every* method above to enter the consensus
(inner join). Per protein, `max_logFC` is the individual method's logFC with
the largest absolute value. The seven methods' p-values are then combined
four classical ways - Fisher's method, Tippett's minimum-p method,
Stouffer's Z-method, and an empirical permutation test (10^5 draws) - each
independently BH-adjusted.

**The catch:** all four classical methods assume the p-values being combined
are *independent* tests. Here they mostly aren't - limma/DEqMS/Wilcoxon/
ROTS/DEP/proDA all run on the same (or closely related) imputed intensity
matrix, so their p-values are correlated. Under positive correlation,
Fisher's method in particular is anti-conservative (inflated false-positive
rate) - this is empirically visible in `Figs/Figure2_ConsensusValidation.pdf`
(panel C: a QQ-plot showing plain Fisher's p-values are systematically too
small relative to a uniform null).

Two corrections are computed alongside the plain combination:

1. **Correlation-corrected Fisher** (`FinalP_FisherCorrected` /
   `FinaladjP_FisherCorrected`, Kost & McDermott, 2002): the average
   pairwise correlation among the methods' $-2\ln p$ statistics is estimated
   once, globally, then used to rescale Fisher's chi-square statistic's
   null distribution per protein.
2. **Vote counting** (`n_methods_tested` / `n_methods_hit`): per protein, how
   many methods it was testable in, and how many of those independently
   clear their own $p \le 0.05$ & $|log_2FC| > 0.25$ threshold. Makes no
   independence assumption at all, at the cost of no single combined
   effect size.

```{r inspect-combined}
cat("Consensus table rows:", nrow(out$combined), "\n")
print(head(out$combined[, c("ID", "max_logFC", "FinalP_Fisher",
                             "FinaladjP_FisherCorrected", "n_methods_hit")]))
```

## Step 5 - Call consensus DEPs (`call_deps`)

`run_pl_dea()` already called DEPs internally (using
`opts$dep_p_col`/`dep_p_cutoff`/`dep_logfc_cutoff`/`dep_min_methods`,
defaulting to correlation-corrected Fisher + logFC + >=3-method agreement)
and wrote them to `out_dir/NegKTSwt_vs_Mock__DEPs.csv`:

```{r deps-default}
out$deps
```

You can also re-call DEPs on the already-computed consensus table with
different cutoffs, without re-running the whole pipeline:

```{r deps-recall}
deps <- call_deps(out$combined, p_col = "FinaladjP_FisherCorrected",
                   logfc_cutoff = 0.25, min_methods = 3)
cat("Consensus DEPs (adj. corrected-Fisher p<=0.05 & |max_logFC|>0.25 & >=3 methods):",
    nrow(deps), "\n")
```

The three cutoffs being combined here: `p_col`/`p_cutoff` (the
correlation-corrected combined p-value, correcting for the seven methods'
non-independence), `logfc_cutoff` (a minimum effect size, independent of
significance), and `min_methods` (requiring at least 3 of the 7 individual
methods to *also* independently agree, rather than trusting the combined
p-value alone) - three different, complementary lines of evidence a protein
has to clear before being called a DEP.

## Where to go next

This walkthrough covers the **intensity branch** only
(`run_pl_dea()`/`call_deps()`), matching `example_NegKTSwt_vs_Mock.R`
exactly. The toolkit also has a **spectral-count branch** (a SAINT-*inspired*
Bayesian mixture model, independent of MaxLFQ intensity entirely) that can
corroborate or add proteins the intensity branch misses - e.g. proteins
identified only via non-unique/shared peptides, where MaxLFQ's cross-run
ratio estimation is unreliable even though raw spectral counts show an
unambiguous, highly reproducible enrichment. See
`Examples/example_fused_NegKTSwt_vs_Mock.R` (`run_pl_interactome()`) for that
dual-branch fusion, and `Figs/Figure3_DualBranchFusion.pdf` for why it
matters on this exact dataset.
